Distributional Trees for Circular Data

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
figure placement 'p' rather than 'ht'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2437 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b0656aaab27d43a3b76cc3c11e51277bf7c1a375
Correct minor errors and update domain of mu
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2438 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 6da81f5f674e38e5ce1f31cdafc30d945ba87dc0

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in \mathbb{R}$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

@@

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates employed are
wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley.}
\label{fig:tree} \end{figure}

@@

As an illustration of the tree-based wind direction forecasts at Innsbruck
Airport, a distributional tree of depth three is depicted in

@@

Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is modeling
wind direction in mountainous terrain where wind shifts can occur due to turns
a turn of the pressure gradient along a valley.

>>>>

In this study distributional trees are adapted to circular responses by employing the von Mises
distribution, also known as ``the circular normal distribution''. Based on a location parameter
$\mu \in [0, 2 \pi]$ and a concentration parameter $\kappa > 0$ the density for $y \in [0, 2 \pi]$
is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
(see, e.g., Jammalamadaka and Sengupta (2001) for a more detailed overview).

@@

\begin{figure}[p!]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates employed are
wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley.}
\label{fig:tree} \end{figure}

@@

@@

Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This allows to model
wind direction in mountainous terrain where wind shifts can occur due to turns
of the pressure gradients along a valley.
